\documentclass[11pt]{article}

\usepackage[a4paper,margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage[hidelinks]{hyperref}
\usepackage{enumitem}

\emergencystretch=4em
\sloppy

\title{Fourth Codex Review Findings for\\
\texttt{rational\_cut\_and\_project\_gap\_periods.tex}}
\author{Codex Review Notes}
\date{May 26, 2026}

\begin{document}

\maketitle

\section*{Executive Summary}

I reviewed the current manuscript
\[
\texttt{LaTeX/rational\_cut\_and\_project\_gap\_periods.tex}
\]
again, including the newly inserted explanatory figures.  I did not find a
mathematical blocker in the rational multiset theorem, the set-valued theorem,
the generic minimality proof, or the degenerate case.  The residue argument
remains coherent:
\[
r=\alpha x-\beta y,\qquad
\tilde p=\beta x+\alpha y,\qquad
c_r\equiv -\alpha\beta^{-1}r \pmod D.
\]

The most important remaining issue is expository rather than structural:
two new schematic figures use values of \(D\) that cannot occur as
\(D=\alpha^2+\beta^2\) for coprime positive integers \(\alpha,\beta\).
A strict referee may notice this.  The figures should therefore either be
explicitly marked as schematic illustrations, or replaced by examples with
realizable \(D\)-values.

\section*{Findings}

\begin{enumerate}[label=\textbf{F\arabic*.},leftmargin=*]

\item \textbf{The new figures use non-realizable \(D\)-values.}

The figure around
\[
\texttt{LaTeX/rational\_cut\_and\_project\_gap\_periods.tex:376}
\]
uses \(D=6\), and the heavy-set figure around
\[
\texttt{LaTeX/rational\_cut\_and\_project\_gap\_periods.tex:911}
\]
uses \(D=12\).  In the actual model, however,
\[
D=\alpha^2+\beta^2
\]
with \(\alpha,\beta\in\mathbb N\) coprime.  Neither \(6\) nor \(12\) is
representable in that form.

\textbf{Recommendation.}  Either mark both figures explicitly as schematic,
for example ``schematic residue picture, not tied to a particular
\((\alpha,\beta)\)'', or replace the examples by realizable values of \(D\).
This is the strongest issue in the current draft because it is easy for a
careful reader to check.

\item \textbf{The helpful \(j\)-clarification is missing again.}

In the block enumeration around
\[
\texttt{LaTeX/rational\_cut\_and\_project\_gap\_periods.tex:799--809},
\]
the text writes
\[
i=kN+j,\qquad 0\le j<N,
\]
and then
\[
\tilde p_s^{(i)}=kD+v_j.
\]
This is correct, but the local-index point is subtle enough that a reader may
wonder why the residue list is indexed by \(j\) rather than by \(i\).

\textbf{Recommendation.}  Restore the short clarification:
\begin{quote}
Here \(j\), not \(i\), indexes the residue list
\(v_0,\ldots,v_{N-1}\), since \(j\) is the position of \(i\) within one
block of \(N\) consecutive entries.
\end{quote}

\item \textbf{Use ``\(m\)-stepping parametrization'' in the summary language.}

The abstract and introduction say that the heavy residue classes form a
proper cyclic interval in the ``natural residue parametrization''.  In the
proof, the precise statement is that they form such an interval after passing
to the \(m\)-stepping parametrization
\[
f(x)=\mu(mx).
\]

\textbf{Recommendation.}  Replace the summary phrase by
``in the \(m\)-stepping parametrization'' or
``after relabeling by the \(m\)-stepping parametrization''.  This avoids the
impression that the heavy set is necessarily contiguous in the ordinary
numeric order modulo \(D\).

\item \textbf{The Lean section uses ``two layers'' in two different senses.}

In the architecture subsection, the ``first layer'' refers to the abstract
Lean typeclass argument, while the ``second layer'' is the concrete Lean
instance.  Later, in the scope subsection, ``Layer~1'' refers to the
hand-verified strip-to-residue geometry, and ``Layer~2'' refers to the
machine-checked residue model.

\textbf{Recommendation.}  Rename one of these decompositions.  For example:
\begin{itemize}[leftmargin=*]
  \item use ``abstract Lean layer'' and ``concrete Lean layer'' in the
  architecture subsection;
  \item reserve ``Layer~1'' and ``Layer~2'' for the scope distinction between
  hand-verified geometry and machine-checked residue combinatorics.
\end{itemize}

\item \textbf{The irrational section uses density before proving it.}

In the irrational preamble, density of
\[
s(\mathbb Z^2)=\{n-am:m,n\in\mathbb Z\}
\]
is used in the unboundedness construction before it is justified later in
Step~3 of the irrational aperiodicity proposition.

\textbf{Recommendation.}  This is not a mathematical error, but it would read
more cleanly if density were stated as a small lemma before the unboundedness
paragraph.  Then both the unboundedness construction and Step~3 could refer
back to that lemma.

\item \textbf{The multiset-versus-set figure should say whether it is
schematic or concrete.}

The first new figure is pedagogically useful: it shows zero gaps in the
multiset convention and their disappearance in the set convention.  However,
because the caption currently says ``here \(D=6\)'', it reads like a concrete
instance of the paper's \((\alpha,\beta,\omega)\)-model.

\textbf{Recommendation.}  If the figure remains schematic, say so explicitly
in the caption.  If it is meant to be a concrete example, choose a realizable
\(D=\alpha^2+\beta^2\).

\end{enumerate}

\section*{Checks Performed}

\begin{itemize}[leftmargin=*]
  \item The paper was compiled with \texttt{latexmk}; the build completed
  successfully.
  \item The searched LaTeX log showed no undefined references, no multiply
  defined labels, no citation warnings, and no overfull boxes.  The remaining
  warnings were harmless underfull boxes in the Lean/table area.
  \item Running
  \[
  \texttt{python3 src/python/verify\_paper.py}
  \]
  reported zero multiset mismatches, zero set mismatches, and zero negative
  set sentinels over all \(51{,}729\) CSV rows.
  \item The Lean source files
  \[
  \texttt{Basic.lean},\qquad \texttt{Irrational.lean}
  \]
  contain \(3758\) lines in total.  A source search found no occurrences of
  \texttt{sorry}, \texttt{admit}, or actual \texttt{axiom} declarations in
  those Lean files.  I did not rerun \texttt{lake build} in this pass.
\end{itemize}

\section*{Overall Assessment}

The paper remains mathematically strong.  The rational proof is especially
clean: the projected multiset is reduced to repeated residue blocks modulo
\(D\), and the minimality proof via the trivial stabilizer of a proper cyclic
interval is convincing.

Before submission, I would fix the schematic-figure issue and restore the
\(j\)-index clarification.  The other points are mostly wording and
organization improvements, but they would make the manuscript harder for a
strict reader to misread.

\end{document}
